A semantics of core AUTOSAR

### 2016-08-23 paper/paper.tex

Some components merely act as containers of other components structures, these are called \textbf{composition} SWCs. In contrast, \textbf{atomic} SWCs define their own behavior in terms of \textbf{runnables} and \textbf{inter-runnable variables}, primarily. A runnable denotes a sequential and (normally) terminating piece of code, although its actual behavioral definition is typically relegated to external C or Matlab files rather than being part of the AUTOSAR model proper. Instead, an AUTOSAR runnable  only provides constraints on what ports and inter-runnable variables the actual implementation can access; as well as \textbf{event} declarations that determine under what conditions it will get executed.

A runnable under execution is called a \textbf{runnable instance}, and these may execute concurrently even if they belong to the same SWC. Each runnable indicates whether or not it may be instantiated concurrently with itself. Runnable code can also use \textbf{exclusive areas} (a form of mutex semaphores) to further control concurrent behavior.

@@

The process terms of our calculus are as follows:
\begin{itemize}
\item $\Trunnable {\Tr r i} t \Act n$  \newline
Dynamic state for runnable $r$ of component $i$, indicating $n$ current instances, $t$ seconds left until next possible instantiation, and activation state $\Act$.
\item $\Trinst {\Tr r i} c \Xs \Code$ \newline
An instance of runnable $r$ of component $i$, currently executing $\Code$ and owning the exclusive areas $\Xs$, possibly running on behalf of client $c$.
\item $\Texcl {\Tx x i} u$ \newline
Exclusive area $x$ of component $i$, with boolean occupation state $u$.
\item $\Tirv {\Ts s i} v$ \newline
Inter-runnable variable $s$ of component $i$, currently holding value $v$.
\item $\Tdelem {\Tep e p i} u v$ \newline
Non-queued sender-receiver data element $e$ in port $p$ of component $i$, currently holding value \V{V} with boolean update status flag $u$.
\item $\Tqelem {\Tep e p i} n \Vs$ \newline
Queued sender-receiver data element $e$ in port $p$ of component $i$ with maximum capacity $n$, currently holding value sequence $\Vs$.
\item $\Toper {\Top o p i} m \Srv$ \newline
Client-server operation $o$ in port $p$ of component $i$, currently holding sequence number $m$ and call state $\Srv$.
\item $\Ttimer {\Tr r i} {t_p} t$ \newline
Periodic timer for runnable $r$ of component $i$, with a period of $t_p$ and $t$ seconds left of its current cycle.
\end{itemize}

In the untyped setting of this report, a value $v$ can be any data item computed and communicated by an AUTOSAR system, including the unit value $\Tvoid$ and the error codes that might be returned from RTE calls. Meta-variables $n$ and $m$ stand for natural numbers, while $u$ and $t$ range over boolean values and (floating point) time values, respectively. The client $c$ of a runnable instance is a tuple holding the address, sequence number and argument value of the current invocation (if the runnable was invoked by a client-server call), or is $\Tvoid$ otherwise. An activation state $\Act$ either toggles between $\Tidle$ and $\Tpending$, or is a sequence of client tuples (in the case of a runnable triggered by client-server calls). A call state $\Srv$ alternates between indicating an ongoing call with a timeout and a finished call with a result value.
Formally,
\[
\begin{array}{rcll}
  v        & \in & \text{Values} \\
  n,m      & \in & \text{Natural numbers} \\
  u        & \in & \text{Boolean values} \\
  t        & \in & \text{Time values} \\
  c        & \in & \text{Clients} \\
  \Act     & \in & \text{Activation states} \\
  \Srv     & \in & \text{Call states} \\ \\
  c        & ::= & \Tclient {\Top o p i} m v \sep \Tvoid \\
  \Act     & ::= & \Tidle \sep \Tpending \sep \Cs \\
  \Srv     & ::= & \Tcalling t \sep \Tdone v
\end{array}
\]
We write $\Vs$ for a sequence of values \V{V}, $\Cs$ for a sequence of clients \V{C}, etc. Sequences may be empty, which is written $\epsilon$. By $\Tlength{\Vs}$ we mean the length of sequence $\Vs$. Left (right) concatenation of an element with a sequence is written $\Tcons{\V{V}}{\V{Vs}}$ ($\Tsnoc{\V{Vs}}{\V{V}}$).

@@

To this end, we have chosen to ignore all internal computations and just capture the observable effects of runnable execution as a pure sequence of RTE (VFB) calls. And because the result from an RTE call may in general affect subsequent runnable behavior, we use a continuation-passing style where every $\Code$ term except the final function return literally contain the sequence that follows it, as a continuation parameter $\Cont$.
%
\[
\begin{array}{lll}
  \Code & \in & \text{Code}                \\
  \Cont & \in & \text{Values} \rightarrow \text{Code}   \\ \\
  \Code & ::= & \Enter x \Cont               \\
        & |   & \Exit x \Cont                \\
        & |   & \IrvWrite s v \Cont          \\
        & |   & \IrvRead s \Cont             \\
        & |   & \Send {\Te e p} v \Cont      \\
        & |   & \Receive {\Te e p} \Cont     \\
        & |   & \Write {\Te e p} v \Cont     \\
        & |   & \Read {\Te e p} \Cont        \\
        & |   & \Invalidate {\Te e p} \Cont  \\
        & |   & \IsUpdated {\Te e p} \Cont   \\
        & |   & \Call {\To o p} v \Cont      \\
        & |   & \Result {\To o p} \Cont      \\
        & |   & \Return v                    \\
\end{array}
\]

@@

The effect of these agreement rules closely resembles the idea of a \emph{broadcast}: what one process says may be heard by every other process in a large parallel composition. This behavior serves us well, because in general, what one AUTOSAR runnable does may have impact on many (if not all) parts of the executing system. It is however important not to confuse this broadcasting notion with any particular form of AUTOSAR communication. The way transition labels distribute over the parallel composition operator is merely a technical aspect of our process calculus, and will be used to express several different AUTOSAR communication semantics, among other things.

@@

Runnables with an activation state distinct from $\Tidle$ or $\Teps$ are amenable to instantiation. Axiom (49) defines this crucial step for runnables triggered by all types of events except server invocation, which is instead handled by axiom (50). Both rules have many details in common:
\begin{itemize}
\item They apply only when the runnable instantiation timer (parameter two) has reached zero.
\item They reset the timer to a model-defined \emph{minimumStartInterval} to disable further instantiation immediately after a new instance is born.
\item They are guarded by the condition that the current number of runnable instances (fourth parameter) is either zero, or the runnable has been defined to accept concurrent invocations.
\item They make the runnable keep track of the new number of instances.
\item They let the new instance inherit the address of its parent runnable.
\item They create the new instance in parallel with its parent (runnable instances is actually the only example of dynamic process creation in this calculus).
\item They initialize the new instance with code according to the model-defined implementation and an empty list of owned exclusive areas.
\end{itemize}
The differences all emanate from the server/non-server distinction:
\begin{itemize}
\item The non-server runnable instance gets a $\Tvoid$ client parameter whereas the server instance is assigned the client tuple from the head of the server's client queue.
\item The non-server runnable toggles back to the $\Tidle$ activation state, while the server just chops of the head of its client queue.
\item Only the server runnable instance is given a non-void code parameter, which is taken from the same client queue head.
\end{itemize}
The transition label of axioms (49) and (59) is noteworthy because it is not matched by a corresponding hearing transition; all other process terms just ignore the information that a new runnable instance has been spawned. An example of process instantiation within a parallel context will thus merely be a trivial variant of one of the instantiation axioms, and is therefore left out.

>>>>

Some components merely act as containers of other components structures, these are called \textbf{composition} SWCs. In contrast, \textbf{atomic} SWCs define their own behaviour in terms of \textbf{runnables} and \textbf{inter-runnable variables}, primarily. A runnable denotes a sequential and (normally) terminating piece of code, although its actual behavioural definition is typically relegated to external C or Matlab files rather than being part of the AUTOSAR model proper. Instead, an AUTOSAR runnable only provides constraints on what ports and inter-runnable variables the actual implementation can access; as well as \textbf{event} declarations that determine under what conditions it will get executed.

A runnable under execution is called a \textbf{runnable instance}, and several instances may execute concurrently even if they belong to the same SWC. Each runnable indicates whether or not it may be instantiated concurrently with itself. Runnable code can also use \textbf{exclusive areas} (a form of mutex semaphores) to further control concurrent behaviour.

@@

The process terms of our calculus are as follows:
\begin{itemize}
\item $\Trunnable {\Tr r i} t \Act n$  \newline
Dynamic state for runnable $r$ of component $i$, indicating $n$ current instances, $t$ seconds left until next possible instantiation, and activation state $\Act$.
\item $\Trinst {\Tr r i} c \Xs \Code$ \newline
An instance of runnable $r$ of component $i$, currently executing $\Code$ and owning the exclusive areas $\Xs$, possibly running on behalf of client $c$.
\item $\Texcl {\Tx x i} u$ \newline
Exclusive area $x$ of component $i$, with boolean occupation state $u$.
\item $\Tirv {\Ts s i} v$ \newline
Inter-runnable variable $s$ of component $i$, currently holding value $v$.
\item $\Tdelem {\Tep e p i} u v$ \newline
Non-queued sender-receiver data element $e$ in port $p$ of component $i$, currently holding value $v$ with boolean update status flag $u$.
\item $\Tqelem {\Tep e p i} n \Vs$ \newline
Queued sender-receiver data element $e$ in port $p$ of component $i$ with maximum capacity $n$, currently holding value sequence $\Vs$.
\item $\Toper {\Top o p i} m \Srv$ \newline
Client-server operation $o$ in port $p$ of component $i$, currently holding sequence number $m$ and call state $\Srv$.
\item $\Ttimer {\Tr r i} {t_p} t$ \newline
Periodic timer for runnable $r$ of component $i$, with a period of $t_p$ and $t$ seconds left of its current cycle.
\end{itemize}

In the untyped setting of this report, a value $v$ can be any data item computed and communicated by an AUTOSAR system, including the unit value $\Tvoid$ and the error codes that might be returned from RTE calls. Meta-variables $n$ and $m$ stand for natural numbers, while $u$ and $t$ range over boolean values and (floating point) time values, respectively. The client $c$ of a runnable instance is a tuple holding the address, sequence number and argument value of the current invocation (if the runnable was invoked by a client-server call), or is $\Tvoid$ otherwise. An activation state $\Act$ either toggles between $\Tidle$ and $\Tpending$, or is a sequence of client tuples (in the case of a runnable triggered by client-server calls). A call state $\Srv$ alternates between indicating an ongoing call with a timeout and a finished call with a result value.
Formally,
\[
\begin{array}{rcll}
  v        & \in & \text{Values} \\
  n,m      & \in & \text{Natural numbers} \\
  u        & \in & \text{Boolean values} \\
  t        & \in & \text{Time values} \\
  c        & \in & \text{Clients} \\
  \Act     & \in & \text{Activation states} \\
  \Srv     & \in & \text{Call states} \\ \\
  c        & ::= & \Tclient {\Top o p i} m v \sep \Tvoid \\
  \Act     & ::= & \Tidle \sep \Tpending \sep \Cs \\
  \Srv     & ::= & \Tcalling t \sep \Tdone v
\end{array}
\]
We write $\Vs$ for a sequence of values \V{V}, $\Cs$ for a sequence of clients \V{C}, etc. Sequences may be empty, which is written~$\epsilon$. By $\Tlength{\Vs}$ we mean the length of sequence $\Vs$. Left (right) concatenation of an element with a sequence is written $\Tcons{\V{V}}{\V{Vs}}$ ($\Tsnoc{\V{Vs}}{\V{V}}$).

@@

To this end, we have chosen to ignore all internal computations and just capture the observable effects of runnable execution as a pure sequence of RTE (VFB) calls. And because the result from an RTE call may in general affect subsequent runnable behaviour, we use a continuation-passing style where every $\Code$ term except the final function return literally contains the sequence that follows it, as a continuation parameter $\Cont$.
%
\[
\begin{array}{lll}
  \Code & \in & \text{Code}                \\
  \Cont & \in & \text{Values} \rightarrow \text{Code}   \\ \\
  \Code & ::= & \Enter x \Cont               \\
        & |   & \Exit x \Cont                \\
        & |   & \IrvWrite s v \Cont          \\
        & |   & \IrvRead s \Cont             \\
        & |   & \Send {\Te e p} v \Cont      \\
        & |   & \Receive {\Te e p} \Cont     \\
        & |   & \Write {\Te e p} v \Cont     \\
        & |   & \Read {\Te e p} \Cont        \\
        & |   & \Invalidate {\Te e p} \Cont  \\
        & |   & \IsUpdated {\Te e p} \Cont   \\
        & |   & \Call {\To o p} v \Cont      \\
        & |   & \Result {\To o p} \Cont      \\
        & |   & \Return v                    \\
\end{array}
\]

@@

The effect of these agreement rules closely resembles the idea of a \emph{broadcast}: what one process says may be heard by every other process in a large parallel composition. This behaviour serves us well, because in general, what one AUTOSAR runnable does may have impact on many (if not all) parts of the executing system. It is however important not to confuse this broadcasting notion with any particular form of AUTOSAR communication. The way transition labels distribute over the parallel composition operator is merely a technical aspect of our process calculus, and will be used to express several different AUTOSAR communication semantics, among other things.

%% TODO: PaJa: continue reading from here.

@@

Runnables with an activation state distinct from $\Tidle$ or $\Teps$ are amenable to instantiation. Axiom (49) defines this crucial step for runnables triggered by all types of events except server invocation, which is instead handled by axiom (50). Both rules have many details in common:
\begin{itemize}
\item They apply only when the runnable instantiation timer (parameter two) has reached zero.
\item They reset the timer to a model-defined \emph{minimumStartInterval} to disable further instantiation immediately after a new instance is born.
\item They are guarded by the condition that the current number of runnable instances (fourth parameter) is either zero, or the runnable has been defined to accept concurrent invocations.
\item They make the runnable keep track of the new number of instances.
\item They let the new instance inherit the address of its parent runnable.
\item They create the new instance in parallel with its parent (runnable instances is actually the only example of dynamic process creation in this calculus).
\item They initialise the new instance with code according to the model-defined implementation and an empty list of owned exclusive areas.
\end{itemize}
The differences all emanate from the server/non-server distinction:
\begin{itemize}
\item The non-server runnable instance gets a $\Tvoid$ client parameter whereas the server instance is assigned the client tuple from the head of the server's client queue.
\item The non-server runnable toggles back to the $\Tidle$ activation state, while the server just chops of the head of its client queue.
\item Only the server runnable instance is given a non-void code parameter, which is taken from the same client queue head.
\end{itemize}
The transition label of axioms (49) and (59) is noteworthy because it is not matched by a corresponding hearing transition; all other process terms just ignore the information that a new runnable instance has been spawned. An example of process instantiation within a parallel context will thus merely be a trivial variant of one of the instantiation axioms, and is therefore left out.
